paper figures

### 2025-01-17 paper_figures.Rmd
capitalization

columns <- list('diversity'=c('nclusters_final_downsampled',

@@

                'isotypes'=c('alpha_ratio_clusters',

@@

                'structural'=c('insert_frequency',

@@

                'context'=c('homology_fw','homology_rv',

@@

                'breaks'=c('mean_length',

@@

                'context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',

@@

                'diversity_raw'=c('nclusters_final',

@@

                'variants'=c('frac_variants_germline',

@@

                                levels=c('library',
                                         'isotypes',
                                         'diversity',
                                         'diversity_raw',
                                         'structural',
                                         'context',
                                         'context_isotype',
                                         'breaks',
                                         'variants')))

cat_cols <- c('structural'='#FD9E40',
              'context'='#FED6AC',
              'context_isotype'='#FED6AC',
              'breaks'='#1E8AC4',
              'isotypes'='#AEF58E',
              'diversity_raw'='#3A7A2C',
              'diversity'='#3A7A2C',
              'variants'='lightpink3',
              'library'='gray40')

@@

tissue_colors <- c('splenocytes'='orange2','lymph node'='orangered3','CH12'='forestgreen')
var_colors <- c('batch'='palegreen2','diagnosis'='tan2','donor'='firebrick4','sequencing'='slategray1','unexplained'='gray80',
                'sex'='steelblue2','age'='darkseagreen4')

diagnosis_colors <- c('control'='gray20','HD'='gray20',
                      'subclass'='lightslategray','CVID'='coral1',
                      'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                      'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
                      'other'='#BC80BD','AID'='darkgoldenrod3',
                      'ATM'='#A65628','LIG4'='mediumpurple2',
                      'NIPBL'='#FCCDE5','BRCA1'='darkkhaki',
                      'NFKB2'='#FFED6F','PI3K'='#80B1D3')

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=nclusters_final, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# clones', y='# filtered clusters', color='# reads')
```

@@

ggplot(cellswitch.data %>%
         dplyr::mutate(donor=plyr::revalue(donor,
                                           c('21084'='A',
                                             '21085'='B',
                                             '21086'='C'))),
       aes(x=ncells, y=nclusters_final)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(aes(color=donor),size=.5) + 
  geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(legend.position=c(.8,.25),
        legend.background = element_rect(fill=NA)) +
  labs(x='# input cells', y='# filtered clusters', color='donor')
```

@@

vdj_stats <- VDJ_results %>%
  dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
  dplyr::group_by(donor,ncells,sample) %>%
  dplyr::summarise(nclusters_final=n_distinct(clone),
                   alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
  dplyr::mutate(method='VDJ')

swibrid_stats <- cellswitch.data %>%
  dplyr::select(donor,ncells,
                nclusters_final,
                alpha_ratio_clusters) %>%
  dplyr::filter(ncells %in% c(50000,100000)) %>%
  dplyr::mutate(method='swibrid')

@@

vdj_df %>% 
  dplyr::filter(metric=='nclusters_final') %>%
  ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_point(aes(color=gsub('donor ','',donor)),size=.5) +
  facet_wrap(~method, ncol=2, scales='free_y') + 
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.clip='off',
        legend.position='none',
        legend.background =element_rect(fill=NA)) +
  labs(y='# clones',x='# cells / 1000',color='')
```

@@

```{r Fig1G, fig.width=1.4 * fig_scale,fig.height=1.4 * fig_scale}
vdj_df %>% 
  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
                donor=gsub('donor ','',donor)) %>%
  dplyr::filter(metric=='alpha_ratio_clusters') %>%
  dplyr::group_by(ncells, donor, method) %>%
  dplyr::summarise(mean_alpha_ratio=mean(value),
                   se_alpha_ratio=sd(value)/sqrt(length(value))) %>%
  pivot_wider(id_cols=c(ncells,donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_swibrid,
             xmin=mean_alpha_ratio_VDJ-se_alpha_ratio_VDJ,
             xmax=mean_alpha_ratio_VDJ+se_alpha_ratio_VDJ,
             ymin=mean_alpha_ratio_swibrid-se_alpha_ratio_swibrid,
             ymax=mean_alpha_ratio_swibrid+se_alpha_ratio_swibrid)) +
  geom_point(aes(color=donor,shape=ncells),size=1) +
  geom_errorbar(linewidth=.25) +
  geom_errorbarh(linewidth=.25) +
  scale_x_continuous(breaks=c(.6,.65)) +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.background=element_blank()) +
  labs(y='fraction IGHA swibrid',x='fraction IGHA VDJ',shape='# cells')
```

@@

HD.scaled <- HD.data %>%
  dplyr::filter(cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var <- sapply(unique(HD.scaled$feature), 
                 function(ftr)
                   tryCatch(my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
                                                      HD.scaled %>%
                                                        dplyr::filter(feature==ftr)),
                            error=setNames(rep(NA,5), c('donor','sex','age','sequencing','batch')))) %>%
                   t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

@@

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=rev(c('donor','sex','age','batch','sequencing','unexplained'))),
                category=factor(category, levels=c('structural','context','breaks','diversity','isotypes'))) %>%
  dplyr::filter(col %in% cols.here) %>%
  dplyr::mutate(col=factor(gsub('_downsampled|_switch','',col), levels=gsub('_downsampled|_switch','',cols.here))) %>%
  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
  geom_col(width=.8) +
  my_theme() +
  facet_wrap(~category,scales='free_y', ncol=5) +
  theme(strip.clip='off',
        legend.position='bottom',
        legend.box='horizontal') +
  guides(fill=guide_legend(ncol=6)) +
  coord_flip() +
  scale_y_continuous(breaks=c(0,50,100)) +
  scale_fill_manual(values=var_colors,
                    breaks=c('donor','sex','age','batch','sequencing','unexplained')) +
  labs(x='', y='% explained variance', fill='')
```

```{r Fig2D, fig.width=1.75 * fig_scale,fig.height=1.7 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('diversity_raw','context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                donor >= donor_cutoff, 
                sequencing < sequencing_cutoff, 
                batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=c('sequencing','batch','donor','sex','age'))) %>%
  dplyr::filter(variable!='unexplained', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=variable, y=100*variance)) + 
  geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
  geom_jitter(aes(color=ifelse(col %in% selected.cols, 'selected','unused')), position=position_jitter(width=.25), size=.25) +
  annotate("rect", xmin = .5, xmax = 1.5, ymax = 100*sequencing_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymax = 100*batch_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 2.5, xmax = 3.5, ymax = Inf, ymin = 100*donor_cutoff, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate('text',.6,78,label='unused',color='black',size=2, hjust=0) +
  annotate('text',.6,73,label='selected',color='orangered',size=2, hjust=0) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='none',
        legend.box='vertical') +
  scale_fill_manual(values=var_colors, guide='none') +
  scale_color_manual(values=c('unused'='black','selected'='orangered')) +
  labs(x='', y='% explained variance', fill='', color='')
```

@@

```{r FigS2F,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=c('donor','sex','age','batch','sequencing','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('library','isotypes','diversity','structural','context','breaks','variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% explained variance', fill='')
```

@@

```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('diversity','diversity_raw'))

@@

ggplot(tmp,aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  guides(fill=guide_legend(title.position='top')) +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('diversity'='blue3','diversity_raw'='lightblue1'),
                    labels=c('downsampled','raw')) +
  labs(x='', y='% explained variance', fill='diversity measures')
```

@@

```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('log10_nreads','# reads',variable),
                                levels=rev(c('donor','cohort','batch','sequencing','# reads','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('context','context_isotype'), variable!='cohort')

@@

ggplot(tmp, aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('context'='darkorchid3','context_isotype'='orchid1'),
                    labels=c('global','per isotype')) +
  labs(x='', y='% explained variance', fill='context measures')
```

@@

HD.scaled.br <- HD.data.br %>%
  dplyr::filter(cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var.br <- sapply(unique(HD.scaled.br$feature), 
                    function(ftr)
                      my_fitExtractVarPartModel(value ~ (1 | donor) + (1 | sex) + (1 | age) + (1 | sequencing) + (1 | batch),
                                                HD.scaled.br %>%
                                                  dplyr::filter(feature==ftr))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

@@

tmp <- rbind(HD.var %>% 
               dplyr::mutate(averaging='clusters'), 
             HD.var.br %>%
               dplyr::mutate(averaging='reads')) %>% 
  dplyr::filter(col %in% fts) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, levels=rev(c('donor','sex','age','batch','sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', 
                category %in% c("context","breaks"),
                variable %in% c("donor", "batch", "sequencing")) %>% 
  dplyr::arrange(col,averaging,variable)

@@

ggplot(tmp, aes(x=variable, y=100*variance, fill=averaging)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.1,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  coord_cartesian(clip='off') +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=c('clusters'='aquamarine4','reads'='darkseagreen1')) +
  labs(x='', y='% explained variance', fill='averages over')
```

@@

```{r Fig2E, fig.width=.9 * fig_scale, fig.height=1.7 * fig_scale}
age.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
  dplyr::slice_max(age, n=1) %>%
  dplyr::pull(col)
age.cols <- 'top_clone_occupancy_downsampled'

sex.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('library','diversity_raw','context_isotype'))) %>%
  dplyr::slice_max(sex, n=1) %>%
  dplyr::pull(col)
sex.cols <- 'spread_SG1'

>>>>

columns <- list('Diversity'=c('nclusters_final_downsampled',

@@

                'Isotypes'=c('alpha_ratio_clusters',

@@

                'Structural'=c('insert_frequency',

@@

                'Context'=c('homology_fw','homology_rv',

@@

                'Breaks'=c('mean_length',

@@

                'Context_isotype'=c('homology_fw_SM_SA','homology_rv_SM_SA',

@@

                'Diversity_raw'=c('nclusters_final',

@@

                'Variants'=c('frac_variants_germline',

@@

                                levels=c('Library',
                                         'Isotypes',
                                         'Diversity',
                                         'Diversity_raw',
                                         'Structural',
                                         'Context',
                                         'Context_isotype',
                                         'Breaks',
                                         'Variants')))

cat_cols <- c('Structural'='#FD9E40',
              'Context'='#FED6AC',
              'Context_isotype'='#FED6AC',
              'Breaks'='#1E8AC4',
              'Isotypes'='#AEF58E',
              'Diversity_raw'='#3A7A2C',
              'Diversity'='#3A7A2C',
              'Variants'='lightpink3',
              'Library'='gray40')

@@

tissue_colors <- c('Splenocytes'='orange2','Lymph Node'='orangered3','CH12'='forestgreen')
var_colors <- c('Batch'='palegreen2','Diagnosis'='tan2','Donor'='firebrick4','Sequencing'='slategray1','unexplained'='gray80',
                'Sex'='steelblue2','Age'='darkseagreen4')

diagnosis_colors <- c('Control'='gray20','HD'='gray20',
                      'Subclass'='lightslategray','CVID'='coral1',
                      'ICOS'='#8DD3C7','Kabuki'='#BEBADA',
                      'IgG4def'='burlywood4','TNFRSF13B'='palegreen4',
                      'Atypical'='gray40',
                      'CVID-like'='#BC80BD','AID'='darkgoldenrod3',
                      'ATM'='#A65628','LIG4'='mediumpurple2',
                      'NIPBL'='#FCCDE5','BRCA1'='darkkhaki',
                      'NFKB2'='#FFED6F','PI3K'='#80B1D3')

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=nclusters_final, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
```

@@

ggplot(cellswitch.data %>%
         dplyr::mutate(Donor=plyr::revalue(Donor,
                                           c('21084'='A',
                                             '21085'='B',
                                             '21086'='C'))),
       aes(x=ncells, y=nclusters_final)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(aes(color=Donor),size=.5) + 
  geom_smooth(size=.5, method='lm', alpha=.2, color='black') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(legend.position=c(.8,.25),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Input cells', y='# Filtered clusters', color='Donor')
```

@@

vdj_stats <- VDJ_results %>%
  dplyr::mutate(clone=paste0(gsub('\\([0-9]*\\)$','',allVHitsWithScore,'_',str_length(aaSeqCDR3)))) %>%
  dplyr::group_by(Donor,ncells,Sample) %>%
  dplyr::summarise(nclusters_final=n_distinct(clone),
                   alpha_ratio_clusters=mean(grepl('IGHA',allCHitsWithScore))) %>%
  dplyr::mutate(method='VDJ')

swibrid_stats <- cellswitch.data %>%
  dplyr::select(Donor,ncells,
                nclusters_final,
                alpha_ratio_clusters) %>%
  dplyr::filter(ncells %in% c(50000,100000)) %>%
  dplyr::mutate(method='SWIBRID')

@@

vdj_df %>% 
  dplyr::filter(metric=='nclusters_final') %>%
  ggplot(aes(x=factor(gsub('000$','',ncells), levels=c('50','100')), y=value)) +
  geom_boxplot(outlier.shape=NA, lwd=.5) +
  geom_point(aes(color=gsub('Donor ','',Donor)),size=.5) +
  facet_wrap(~method, ncol=2, scales='free_y') + 
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.clip='off',
        legend.position='none',
        legend.background =element_rect(fill=NA)) +
  labs(y='# Clones',x='# Cells / 1000',color='')
```

@@

```{r Fig1G, fig.width=1.4 * fig_scale,fig.height=1.4 * fig_scale}
vdj_df %>% 
  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
                Donor=gsub('Donor ','',Donor)) %>%
  dplyr::filter(metric=='alpha_ratio_clusters') %>%
  dplyr::group_by(ncells, Donor, method) %>%
  dplyr::summarise(mean_alpha_ratio=mean(value),
                   se_alpha_ratio=sd(value)/sqrt(length(value))) %>%
  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_alpha_ratio','se_alpha_ratio')) %>%
  ggplot(aes(x=mean_alpha_ratio_VDJ, y=mean_alpha_ratio_SWIBRID,
             xmin=mean_alpha_ratio_VDJ-se_alpha_ratio_VDJ,
             xmax=mean_alpha_ratio_VDJ+se_alpha_ratio_VDJ,
             ymin=mean_alpha_ratio_SWIBRID-se_alpha_ratio_SWIBRID,
             ymax=mean_alpha_ratio_SWIBRID+se_alpha_ratio_SWIBRID)) +
  geom_point(aes(color=Donor,shape=ncells),size=1) +
  geom_errorbar(linewidth=.25) +
  geom_errorbarh(linewidth=.25) +
  scale_x_continuous(breaks=c(.6,.65)) +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.background=element_blank()) +
  labs(y='Fraction IGHA SWIBRID',x='Fraction IGHA VDJ',shape='# cells')
```

@@

HD.scaled <- HD.data %>%
  dplyr::filter(Cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var <- sapply(unique(HD.scaled$feature), 
                 function(ftr)
                   tryCatch(my_fitExtractVarPartModel(value ~ (1 | Donor) + (1 | Sex) + (1 | Age) + (1 | Sequencing) + (1 | Batch),
                                                      HD.scaled %>%
                                                        dplyr::filter(feature==ftr)),
                            error=setNames(rep(NA,5), c('Donor','Sex','Age','Sequencing','Batch')))) %>%
                   t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

@@

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','unexplained'))),
                category=factor(category, levels=c('Structural','Context','Breaks','Diversity','Isotypes'))) %>%
  dplyr::filter(col %in% cols.here) %>%
  dplyr::mutate(col=factor(gsub('_downsampled|_switch','',col), levels=gsub('_downsampled|_switch','',cols.here))) %>%
  ggplot(aes(x=col, y=100*variance, fill=variable)) + 
  geom_col(width=.8) +
  my_theme() +
  facet_wrap(~category,scales='free_y', ncol=5) +
  theme(strip.clip='off',
        legend.position='bottom',
        legend.box='horizontal') +
  guides(fill=guide_legend(ncol=6)) +
  coord_flip() +
  scale_y_continuous(breaks=c(0,50,100)) +
  scale_fill_manual(values=var_colors,
                    breaks=c('Donor','Sex','Age','Batch','Sequencing','unexplained')) +
  labs(x='', y='% Explained variance', fill='')
```

```{r Fig2D, fig.width=1.75 * fig_scale,fig.height=1.7 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                Donor >= donor_cutoff, 
                Sequencing < sequencing_cutoff, 
                Batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

HD.var %>%  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(gsub('Residuals','unexplained',variable),
                                levels=c('Sequencing','Batch','Donor','Sex','Age'))) %>%
  dplyr::filter(variable!='unexplained', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
  ggplot(aes(x=variable, y=100*variance)) + 
  geom_boxplot(aes(fill=variable), outlier.shape=NA, lwd=.5, legend=FALSE) + 
  geom_jitter(aes(color=ifelse(col %in% selected.cols, 'selected','unused')), position=position_jitter(width=.25), size=.25) +
  annotate("rect", xmin = .5, xmax = 1.5, ymax = 100*sequencing_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 1.5, xmax = 2.5, ymax = 100*batch_cutoff, ymin = 0, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate("rect", xmin = 2.5, xmax = 3.5, ymax = Inf, ymin = 100*donor_cutoff, fill='orangered', linetype=1, size=.25, alpha=.25) +
  annotate('text',.6,78,label='unused',color='black',size=2, hjust=0) +
  annotate('text',.6,73,label='selected',color='orangered',size=2, hjust=0) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='none',
        legend.box='vertical') +
  scale_fill_manual(values=var_colors, guide='none') +
  scale_color_manual(values=c('unused'='black','selected'='orangered')) +
  labs(x='', y='% Explained variance', fill='', color='')
```

@@

```{r FigS2F,fig.width=3.5 * fig_scale,fig.height=1.7 * fig_scale}
HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, 
                                levels=c('Donor','Sex','Age','Batch','Sequencing','Residuals'))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Isotypes','Diversity','Structural','Context','Breaks','Variants')) %>%
  ggplot(aes(x=category, y=100*variance, fill=variable)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.25) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=var_colors) +
  labs(x='', y='% Explained variance', fill='')
```

@@

```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Diversity','Diversity_raw'))

@@

ggplot(tmp,aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  guides(fill=guide_legend(title.position='top')) +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('Diversity'='blue3','Diversity_raw'='lightblue1'),
                    labels=c('downsampled','raw')) +
  labs(x='', y='% Explained variance', fill='Diversity measures')
```

@@

```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Context','Context_isotype'))

@@

ggplot(tmp, aes(x=variable, y=100*variance, fill=category)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.3,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        strip.clip='off',
        legend.position='top',
        legend.box='horizontal') +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=c('Context'='darkorchid3','Context_isotype'='orchid1'),
                    labels=c('global','per isotype')) +
  labs(x='', y='% Explained variance', fill='Context measures')
```

@@

HD.scaled.br <- HD.data.br %>%
  dplyr::filter(Cohort=='C1') %>%
  dplyr::group_by(feature) %>%
  dplyr::mutate(value=as.vector(scale(value))) %>%
  dplyr::filter(!is.na(value)) %>%
  dplyr::ungroup()

HD.var.br <- sapply(unique(HD.scaled.br$feature), 
                    function(ftr)
                      my_fitExtractVarPartModel(value ~ (1 | Donor) + (1 | Sex) + (1 | Age) + (1 | Sequencing) + (1 | Batch),
                                                HD.scaled.br %>%
                                                  dplyr::filter(feature==ftr))) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column('col') %>%
  gather(variable,variance,-col)

@@

tmp <- rbind(HD.var %>% 
               dplyr::mutate(averaging='Clusters'), 
             HD.var.br %>%
               dplyr::mutate(averaging='Reads')) %>% 
  dplyr::filter(col %in% fts) %>%
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable, 
                                levels=rev(c('Donor','Sex','Age','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', 
                category %in% c("Context","Breaks"),
                variable %in% c("Donor", "Batch", "Sequencing")) %>% 
  dplyr::arrange(col,averaging,variable)

@@

ggplot(tmp, aes(x=variable, y=100*variance, fill=averaging)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA, lwd=.5) + 
  geom_jitter(position=position_jitterdodge(jitter.width=.1,dodge.width=.8), size=.5) +
  geom_text(data=pvals, aes(x=variable, label=stars.pval(p)), y=Inf, inherit.aes=FALSE, size=5) +
  my_theme() +
  coord_cartesian(clip='off') +
  guides(fill=guide_legend(title.position='top')) +
  theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1),
        legend.position='top',
        legend.box='horizontal') +
  scale_fill_manual(values=c('Clusters'='aquamarine4','Reads'='darkseagreen1')) +
  labs(x='', y='% Explained variance', fill='Averages over')
```

@@

```{r Fig2E, fig.width=.9 * fig_scale, fig.height=1.7 * fig_scale}
age.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
  dplyr::slice_max(Age, n=1) %>%
  dplyr::pull(col)
age.cols <- 'top_clone_occupancy_downsampled'

sex.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Library','Diversity_raw','Context_isotype'))) %>%
  dplyr::slice_max(Sex, n=1) %>%
  dplyr::pull(col)
sex.cols <- 'spread_SG1'
